---
title: "Week 9 Coal Data Consumption"
author: "Sai Sudheer Dontha"
date: "2023-11-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Reading the dataset
```{r}
coalConsumptionData <- read.csv("C:/Users/sudhi/Downloads/BIS_581/CoalConsumption_R_Project/annual-coal-consumption-by-country-1980-2009-2.csv")
head(coalConsumptionData, n=8)

```

### Rename the first column to "Region"
```{r}
colnames(coalConsumptionData)[1] <- "Region"
head(coalConsumptionData, n=8)
```
### Display summary statistics of the dataset
```{r}
summary(coalConsumptionData)
```

### Loading the required libraries
### Load the library for pivot_longer and pivot_wider functions.
```{r}
library(tidyverse)  
library(dplyr)
```
### Transforming the dataset into a long format using pivot_longer
```{r}
coalLongData <- pivot_longer(coalConsumptionData, !Region, names_to = "Year", values_to = "Consumption")
coalLongData
```

### Observing mixed data types in the 'Year' column
### Utilizing transform and gsub to remove "X" from the 'Year' column
```{r}
coalLongData <- transform(coalLongData, Year = gsub("X", "", Year))
head(coalLongData, n=8)
```
View(coalLongData)

### Converting 'Year' to a numeric format
```{r}
coalLongData$Year <- as.numeric(as.character(coalLongData$Year))
is.numeric(coalLongData$Year)
```


### Converting 'Consumption' from character to numeric
### Checking if 'Consumption' is numeric
```{r}
is.numeric(coalLongData$Consumption)
```

### Checking if 'Consumption' is in character format
```{r}
is.character(coalLongData$Consumption)
```


### Converting 'Consumption' to numeric
```{r}
coalLongData$Consumption <- as.numeric(coalLongData$Consumption)
is.numeric(coalLongData$Consumption)
summary(coalLongData)
#View(coalLongData)
```

### Reviewing the class types for all columns
```{r}
sapply(coalLongData, class)
summary(coalLongData)
```

### Removing NA values
```{r}
#install.packages("janitor")
#library(janitor)
head(coalLongData, n=10)
```

### Dropping NA
#```{r}
#library(tidyr)
#head(coalLongData,n=10)
#  drop_na()
#  head(coalLongData, n=10)
```

```{r}
sum(is.na(coalLongData))
summary(coalLongData)
```

```{r}
processedCoalData <- coalLongData
```


### Separating continents, regions, and countries for further analysis
```{r}
continentLabels <- c("Africa", "Asia", "Europe", "North America", "Central & South America", "Former U.S.S.R.",
                     "Middle East", "Central African Republic", "Asia & Oceania", "Antarctica")
```

### Using filtering to categorize
```{r}
continentData <- filter(processedCoalData, Region %in% continentLabels)
head(continentData, n=10)

```


View(continentData)

```{r}
otherRegionData <- processedCoalData %>%
  filter(!(Region %in% continentLabels))
head(otherRegionData, n=10)
```

```{r}
countrySpecificData <- otherRegionData[!(otherRegionData$Region %in% "World"),]
head(countrySpecificData, n=10)
```



### 
```{r}
library(ggplot2)

ggplot(continentData, aes(x = as.numeric(Year), y = Consumption, color = Region)) +
  geom_line(size = 1.2, alpha = 0.8, linetype = "solid") +  
  labs(
    title = "Coal Consumption Over Years by Region",
    x = "Year",
    y = "Consumption"
  ) +
  theme_minimal()
```

### Faceted histograms by region
```{r}
ggplot(continentData, aes(x = Consumption)) +
  geom_histogram(bins = 10) +
  facet_wrap(~Region) +
  labs(
    title = "Distribution of Coal Consumption by Region",
    x = "Consumption",
    y = "Frequency"
  ) +
  theme_minimal()
```


### Stacked bar chart of total consumption by year
```{r}
totalConsumptionByYear <- aggregate(Consumption ~ Year + Region, data = continentData, FUN = sum)

ggplot(totalConsumptionByYear, aes(x = as.factor(Year), y = Consumption, fill = Region)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Total Coal Consumption by Year",
    x = "Year",
    y = "Total Consumption"
  ) +
  theme_minimal()
```


### Violin Plot
```{r}
ggplot(continentData, aes(x = as.numeric(Year), y = Consumption, fill = as.factor(Year))) +
  geom_violin(trim = FALSE) +
  labs(
    title = "Violin Plot of Coal Consumption by Year",
    x = "Year",
    y = "Consumption"
  ) +
  theme_minimal()
```


### Scatter plot
```{r}
ggplot(continentData, aes(x = as.numeric(Year), y = Consumption, color = Region)) +
  geom_point() +
  labs(
    title = "Scatter Plot of Coal Consumption Over Years",
    x = "Year",
    y = "Consumption"
  ) +
  theme_minimal()
```


### Aggregate total consumption by region
```{r}
library(ggplot2)
library(dplyr)

totalConsumptionByRegion <- continentData %>%
  group_by(Region) %>%
  summarise(TotalConsumption = sum(Consumption)) %>%
  arrange(desc(TotalConsumption))

### Create bar chart
ggplot(totalConsumptionByRegion, aes(x = reorder(Region, -TotalConsumption), y = TotalConsumption)) +
  geom_bar(stat = "identity", fill = "violet") +
  labs(
    title = "Total Consumption by Region",
    x = "Region",
    y = "Total Consumption"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



 
